Reanalysis Step 2: Multiple Imputation

### 2025-08-30 reanalysis-2-multiple_imputation.qmd
retry

---
title: "Reanalysis Step 2: Multiple Imputation (Clean Version)"
subtitle: "Simplified imputation using mice with clustering"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

```{r setup}
#| include: false
source("dependencies.R")
source("functions.R")
seed <- 2025
setup_analysis(seed = seed)

library(mice)
library(miceadds)
library(VIM)
library(naniar)
library(dplyr)
library(tidyr)
library(knitr)
library(lme4)

# Initialize logging
log_file <- init_log("multiple_imputation_clean")

# ANALYSIS MODE
analysis_mode <- "final"  # "test" or "final"

if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 3
} else {
  n_imputations <- 30
  n_iterations <- 20
}

log_message(paste0("Analysis mode: ", analysis_mode))
log_message(paste0("Imputations: ", n_imputations, ", Iterations: ", n_iterations))
```

## Load Data

```{r load-data}
# Load the prepared dataset
df_original <- readRDS(file.path(reanalysis_data_dir, "df_for_imputation.rds"))
log_message(paste0("Loaded ", nrow(df_original), " observations"))

# Get unique patient count
n_patients <- n_distinct(df_original$e_mrn_deidentified)
log_message(paste0("Number of unique patients: ", n_patients))
```

@@

```{r define-variables}
# Variables that should NOT be imputed and should NOT be used as predictors
vars_no_impute_no_predict <- c(
  "row_id"  # Internal row identifier
)

# Variables that should NOT be imputed but CAN be used as predictors  
vars_no_impute_yes_predict <- c(
  "ever_lapse_binary",  # Exposure - complete
  "cohort_id"           # Data indicator - complete
)

# Cluster variable (patient ID)
cluster_var <- "e_mrn_deidentified"

# Variables to impute - continuous (use PMM with clustering)
vars_impute_continuous <- c(
  "baseline_VA_logMAR",
  "intermediate_VA_logMAR", 
  "outcome_VA_logMAR",
  "age",
  "CCI",
  "DCSI"
)

# Variables to impute - binary
vars_impute_binary <- c(
  "catsurg_before_hitplus_cat"
)

# Variables to impute - categorical
vars_impute_categorical <- c(
  "gender_cat",
  "race_ethnic_cat",
  "insurance_cat"
)

# Variables to impute - ordinal
vars_impute_ordinal <- c(
  "person_dr"  # 0=No_DR, 1=NPDR, 2=PDR
)

# Treatment variables (binary, likely complete)
vars_treatment <- c(
  "other_inject",
  "anti_VEGF",
  "focal_laser_flag",
  "PRP_flag"
)

# Comorbidity indicators (binary factors from step 1)
vars_comorbidity <- c(
  "glaucoma_bef_hitplus_cat",
  "otherretina_bef_hitplus_cat"
)

# All variables for imputation dataset
all_vars <- c(
  cluster_var,
  vars_no_impute_no_predict,
  vars_no_impute_yes_predict,
  vars_impute_continuous,
  vars_impute_binary,
  vars_impute_categorical,
  vars_impute_ordinal,
  vars_treatment,
  vars_comorbidity
)
```

## Prepare Data for MICE

```{r prepare-data}
# Select variables and clean up (rely on step 1 outputs)
df_mice <- df_original |>
  select(any_of(all_vars)) |>
  mutate(
    # Ensure person_dr is ordered factor
    person_dr = if (is.numeric(person_dr)) {
      factor(person_dr, levels = c(0L, 1L, 2L), labels = c("No_DR", "NPDR", "PDR"), ordered = TRUE)
    } else {
      factor(person_dr, levels = c("No_DR", "NPDR", "PDR"), ordered = TRUE)
    },
    # Clean factor levels (remove spaces for formula compatibility)
    across(where(is.character), ~ factor(gsub("[^[:alnum:]_]", "_", .x)))
  ) |>
  dplyr::group_by(e_mrn_deidentified) |>
  dplyr::mutate(
    n_eyes = dplyr::n(),
    eye_index = dplyr::row_number(),
    fellow_baseline_VA = dplyr::if_else(n_eyes == 2, baseline_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_intermediate_VA = dplyr::if_else(n_eyes == 2, intermediate_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_outcome_VA = dplyr::if_else(n_eyes == 2, outcome_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_dr = dplyr::if_else(n_eyes == 2,
                               person_dr[3 - eye_index],
                               factor(NA, levels = levels(person_dr), ordered = TRUE))
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-eye_index) |>
  dplyr::mutate(
    # Convert cluster variable to integer for 2l methods
    e_mrn_deidentified = as.integer(as.factor(e_mrn_deidentified))
  )

# Check missing data pattern
log_message("Missing data summary:")
missing_summary <- df_mice |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))

print(missing_summary)
```

## Setup Imputation Methods

```{r setup-methods}
# Initialize mice to get default setup
ini <- mice(df_mice, maxit = 0, printFlag = FALSE)

# Set up methods vector
methods <- ini$method

# No imputation for these
methods[intersect(vars_no_impute_no_predict, names(methods))] <- ""
methods[intersect(vars_no_impute_yes_predict, names(methods))] <- ""
if (cluster_var %in% names(methods)) {
  methods[cluster_var] <- ""
}

# Continuous variables - use standard PMM (no multilevel)
for (var in vars_impute_continuous) {
  if (var %in% names(methods)) {
    methods[var] <- "pmm"
  }
}

# Binary variables
for (var in vars_impute_binary) {
  if (var %in% names(methods)) {
    methods[var] <- "logreg"
  }
}

# Categorical variables  
for (var in vars_impute_categorical) {
  if (var %in% names(methods)) {
    methods[var] <- "polyreg"
  }
}

# Ordinal variables
for (var in vars_impute_ordinal) {
  if (var %in% names(methods)) {
    methods[var] <- "polr"
  }
}

# Treatment and comorbidity variables (likely complete, but set methods just in case)
for (var in c(vars_treatment, vars_comorbidity)) {
  if (var %in% names(methods)) {
    methods[var] <- "logreg"
  }
}

# Fellow-eye variables: predictors only (do not impute)
vars_fellow <- c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")
for (var in vars_fellow) {
  if (var %in% names(methods)) {
    methods[var] <- ""
  }
}

log_message("Imputation methods configured")
```

## Setup Predictor Matrix

```{r setup-predictors}
# Create predictor matrix based on theoretical relationships with strict temporal ordering
# Start with empty matrix (no predictors)
pred_matrix <- matrix(0, nrow = ncol(df_mice), ncol = ncol(df_mice))
colnames(pred_matrix) <- names(df_mice)
rownames(pred_matrix) <- names(df_mice)

# STEP 1: Ensure complete exclusions for non-imputed variables
# e_mrn_deidentified and row_id should be completely excluded (rows and columns = 0)
if ("e_mrn_deidentified" %in% names(df_mice)) {
  pred_matrix[, "e_mrn_deidentified"] <- 0
  pred_matrix["e_mrn_deidentified", ] <- 0
}
if ("row_id" %in% names(df_mice)) {
  pred_matrix[, "row_id"] <- 0
  pred_matrix["row_id", ] <- 0
}

# Ensure ever_lapse_binary and cohort_id are never imputed (rows = 0) but can be predictors
if ("ever_lapse_binary" %in% rownames(pred_matrix)) {
  pred_matrix["ever_lapse_binary", ] <- 0  # Never imputed
}
if ("cohort_id" %in% rownames(pred_matrix)) {
  pred_matrix["cohort_id", ] <- 0  # Never imputed
}

# Ensure all fellow variables are never imputed (rows = 0)
fellow_vars <- c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")
for (v in fellow_vars) {
  if (v %in% rownames(pred_matrix)) {
    pred_matrix[v, ] <- 0  # Never imputed
  }
}

# Define variable groups for clarity
demographic_vars <- c("age", "gender_cat", "race_ethnic_cat", "insurance_cat")
global_comorbidity_vars <- c("CCI", "DCSI")
ocular_comorbidity_vars <- c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat", "catsurg_before_hitplus_cat")
treatment_vars <- c("anti_VEGF", "focal_laser_flag", "PRP_flag", "other_inject")

# STEP 2: Configure baseline_VA_logMAR predictors
# Predictors: demographics, global comorbidities, ocular comorbidities, exposure, fellow_baseline_VA
# NO treatments, NO intermediate/outcome VAs
if ("baseline_VA_logMAR" %in% rownames(pred_matrix)) {
  # Demographics
  for (v in demographic_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 1
  }
  # Global comorbidities
  for (v in global_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 1
  }
  # Ocular comorbidities and cataract
  for (v in ocular_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 1
  }
  # Exposure
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "ever_lapse_binary"] <- 1
  # Fellow eye baseline only
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "fellow_baseline_VA"] <- 1

  # Explicitly exclude treatments and future timepoints
  for (v in treatment_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 0
  }
  if ("intermediate_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "intermediate_VA_logMAR"] <- 0
  if ("outcome_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "outcome_VA_logMAR"] <- 0
  if ("fellow_intermediate_VA" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "fellow_intermediate_VA"] <- 0
  if ("fellow_outcome_VA" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "fellow_outcome_VA"] <- 0
}

# STEP 3: Configure intermediate_VA_logMAR predictors
# Predictors: baseline_VA, demographics, CCI/DCSI, ocular comorbidities, exposure, fellow_baseline_VA
# NO treatments, NO fellow_intermediate_VA, NO outcome VAs
if ("intermediate_VA_logMAR" %in% rownames(pred_matrix)) {
  # Baseline VA
  if ("baseline_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "baseline_VA_logMAR"] <- 1
  # Demographics
  for (v in demographic_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 1
  }
  # Global comorbidities
  for (v in global_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 1
  }
  # Ocular comorbidities and cataract
  for (v in ocular_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 1
  }
  # Exposure
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "ever_lapse_binary"] <- 1
  # Fellow eye baseline only (NOT intermediate)
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "fellow_baseline_VA"] <- 1

  # Explicitly exclude treatments, fellow_intermediate_VA, and outcome VAs
  for (v in treatment_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 0
  }
  if ("fellow_intermediate_VA" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "fellow_intermediate_VA"] <- 0
  if ("outcome_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "outcome_VA_logMAR"] <- 0
  if ("fellow_outcome_VA" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "fellow_outcome_VA"] <- 0
}

# STEP 4: Configure outcome_VA_logMAR predictors
# Predictors: baseline_VA, intermediate_VA, demographics, CCI/DCSI, ocular comorbidities, exposure, treatments
# Fellow predictors: fellow_baseline_VA and fellow_outcome_VA (NOT fellow_intermediate_VA)
if ("outcome_VA_logMAR" %in% rownames(pred_matrix)) {
  # Prior timepoints
  if ("baseline_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "baseline_VA_logMAR"] <- 1
  if ("intermediate_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "intermediate_VA_logMAR"] <- 1
  # Demographics
  for (v in demographic_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  }
  # Global comorbidities
  for (v in global_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  }
  # Ocular comorbidities and cataract
  for (v in ocular_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  }
  # Exposure
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "ever_lapse_binary"] <- 1
  # Treatments (allowed for outcome)
  for (v in treatment_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  }
  # Fellow predictors: baseline and outcome only
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "fellow_baseline_VA"] <- 1
  if ("fellow_outcome_VA" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "fellow_outcome_VA"] <- 1

  # Explicitly exclude fellow_intermediate_VA
  if ("fellow_intermediate_VA" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "fellow_intermediate_VA"] <- 0
}

# STEP 5: Configure person_dr predictors
# Predictors: demographics, CCI/DCSI, ocular comorbidities, baseline_VA, exposure, treatments, fellow_baseline_VA
# NO fellow_intermediate_VA, NO outcome_VA (to remain strictly pre-outcome)
if ("person_dr" %in% rownames(pred_matrix)) {
  # Demographics
  for (v in demographic_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  }
  # Global comorbidities
  for (v in global_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  }
  # Ocular comorbidities and cataract
  for (v in ocular_comorbidity_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  }
  # Baseline VA
  if ("baseline_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["person_dr", "baseline_VA_logMAR"] <- 1
  # Exposure
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["person_dr", "ever_lapse_binary"] <- 1
  # Treatments
  for (v in treatment_vars) {
    if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  }
  # Fellow predictors: baseline only and fellow_dr
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["person_dr", "fellow_baseline_VA"] <- 1
  if ("fellow_dr" %in% colnames(pred_matrix)) pred_matrix["person_dr", "fellow_dr"] <- 1

  # Explicitly exclude fellow_intermediate_VA and outcome_VA
  if ("fellow_intermediate_VA" %in% colnames(pred_matrix)) pred_matrix["person_dr", "fellow_intermediate_VA"] <- 0
  if ("outcome_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["person_dr", "outcome_VA_logMAR"] <- 0
  if ("fellow_outcome_VA" %in% colnames(pred_matrix)) pred_matrix["person_dr", "fellow_outcome_VA"] <- 0
}

# STEP 6: Configure other variables (demographics, comorbidities, treatments)
# Allow demographics to predict each other
for (v1 in demographic_vars) {
  for (v2 in demographic_vars) {
    if (v1 != v2 && v1 %in% rownames(pred_matrix) && v2 %in% colnames(pred_matrix)) {
      pred_matrix[v1, v2] <- 1
    }
  }
}

>>>>

---
title: "Reanalysis Step 2: Multiple Imputation"
subtitle: "Imputing missing values using mice with strict temporal and fellow-eye constraints"
author: "Erik Westlund"
date: "`r Sys.Date()`"
format: html
---

@@

```{r setup}
#| include: false
source("dependencies.R")
source("functions.R")  # Load formula definitions and helper functions
seed <- 2025
setup_analysis(seed = seed)

# Packages
library(mice)
library(VIM)
library(naniar)
library(dplyr)
library(tidyr)
library(knitr)
library(tibble)

# ANALYSIS MODE: "test" or "final"
# test = 2 imputations, 3 iterations (fast)
# final = 30 imputations, 20 iterations (stability achieved ~20)
analysis_mode <- "test"

if (analysis_mode == "test") {
  n_imputations <- 2
  n_iterations <- 3
} else if (analysis_mode == "final") {
  n_imputations <- 30
  n_iterations <- 20
} else {
  stop("Invalid analysis_mode. Must be 'test' or 'final'")
}

# Initialize logging
log_file <- init_log("multiple_imputation")
log_message(paste0("Analysis mode: ", analysis_mode))
log_message(paste0("Imputations: ", n_imputations, ", Iterations: ", n_iterations))
```

## Load Prepared Data

```{r load-data}
# Load the prepared dataset from Step 1
df_for_imputation <- readRDS(file.path(reanalysis_data_dir, "df_for_imputation.rds"))
variable_lists <- readRDS(file.path(reanalysis_data_dir, "variable_lists.rds"))

ps_model_vars <- variable_lists$ps_model_vars
outcome_model_vars <- variable_lists$outcome_model_vars

# Quick record/patient counts
log_message(paste0("Loaded records: ", nrow(df_for_imputation)))
if ("e_mrn_deidentified" %in% names(df_for_imputation)) {
  log_message(paste0("Unique patients: ", dplyr::n_distinct(df_for_imputation$e_mrn_deidentified)))
}
```

@@

```{r define-imputation-vars}
# Variables to include in imputation model (SOURCE variables only)
# Derived variables will be recreated post-imputation

imputation_vars <- c(
  "e_mrn_deidentified",

  # Temporal VA variables
  "cohort_id",                 # Data indicator only - not used as predictor
  "baseline_VA_logMAR",        # Baseline VA
  "intermediate_VA_logMAR",    # Intermediate follow-up VA (for excluded cohort)
  "outcome_VA_logMAR",         # 2-year outcome (true for submitted, missing for excluded)
  # NOT outcome_va_vi_binary - derived from logMAR threshold

  # Exposure
  "ever_lapse_binary",

  # Demographics/Clinical
  "gender_cat",
  "race_ethnic_cat",
  "insurance_cat",
  "age",                        # Continuous age (source)
  # NOT age_cat - derived from cut(age)
  "CCI",
  "DCSI",

  # Treatment indicators - ALL SOURCES
  "other_inject",
  "anti_VEGF",
  "focal_laser_flag",
  "PRP_flag",
  # NOT treatment_type or any_treatment - derived from these 4

  # Comorbidities (already recoded to 0/1)
  "glaucoma_bef_hitplus_cat",
  "otherretina_bef_hitplus_cat",
  "catsurg_before_hitplus_cat",

  # DR severity - use person_dr (created from No_DR/NPDR/PDR)
  "person_dr"                  # Numeric 0/1/2 → ordered factor
  # NOT dr_severity/person_dr_severity - derived from person_dr
)

df_mice <- df_for_imputation |>
  mutate(
    # Create a unique row identifier to preserve exact row matching
    row_id = dplyr::row_number()
  ) |>
  dplyr::select(row_id, dplyr::all_of(imputation_vars)) |>
  mutate(
    # Convert person_dr to ordered factor for polr imputation method
    person_dr = factor(person_dr, levels = c(0, 1, 2),
                       labels = c("No_DR", "NPDR", "PDR"),
                       ordered = TRUE),
    # Lock factor levels for categorical variables to ensure consistency
    cohort_id = factor(cohort_id),
    gender_cat = factor(gender_cat),
    race_ethnic_cat = factor(race_ethnic_cat),
    insurance_cat = factor(insurance_cat)
  ) |>
  # Fellow-eye variables (computed within patient clusters)
  dplyr::group_by(e_mrn_deidentified) |>
  dplyr::mutate(
    n_eyes = dplyr::n(),
    eye_index = dplyr::row_number(),
    fellow_baseline_VA = dplyr::if_else(n_eyes == 2, baseline_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_intermediate_VA = dplyr::if_else(n_eyes == 2, intermediate_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_outcome_VA = dplyr::if_else(n_eyes == 2, outcome_VA_logMAR[3 - eye_index], as.numeric(NA)),
    fellow_dr = dplyr::if_else(n_eyes == 2,
                               person_dr[3 - eye_index],
                               factor(NA, levels = levels(person_dr), ordered = TRUE))
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-eye_index)
```

## Setup Imputation Model (dry run and methods)

```{r setup-mice}
# Dry run to inspect missingness and defaults
mice_dry <- mice(df_mice, maxit = 0, printFlag = FALSE)

# Missingness summary
miss_vars <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  Pct_Missing = round(mice_dry$nmis/nrow(df_mice) * 100, 1)
) |>
  dplyr::filter(N_Missing > 0) |>
  dplyr::arrange(dplyr::desc(N_Missing))

# Configure methods explicitly
custom_methods <- mice_dry$method

# Temporal VA variables
custom_methods["baseline_VA_logMAR"] <- "pmm"
custom_methods["intermediate_VA_logMAR"] <- "pmm"
custom_methods["outcome_VA_logMAR"] <- "pmm"
custom_methods["cohort_id"] <- ""  # never impute

# Other continuous
custom_methods["CCI"] <- "pmm"
custom_methods["DCSI"] <- "pmm"
custom_methods["age"] <- "pmm"

# Ordinal
custom_methods["person_dr"] <- "polr"

# Binary
custom_methods["catsurg_before_hitplus_cat"] <- "logreg"
custom_methods["ever_lapse_binary"] <- "logreg"  # likely complete, ok to set type
custom_methods["glaucoma_bef_hitplus_cat"] <- "logreg"
custom_methods["otherretina_bef_hitplus_cat"] <- "logreg"
custom_methods["gender_cat"] <- "logreg"  # if binary
custom_methods["other_inject"] <- "logreg"
custom_methods["anti_VEGF"] <- "logreg"
custom_methods["focal_laser_flag"] <- "logreg"
custom_methods["PRP_flag"] <- "logreg"

# Polytomous
custom_methods["race_ethnic_cat"] <- "polyreg"
custom_methods["insurance_cat"] <- "polyreg"

# IDs and fellow-eye variables: never impute
custom_methods["row_id"] <- ""
custom_methods["e_mrn_deidentified"] <- ""
for (fv in c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")) {
  if (fv %in% names(custom_methods)) custom_methods[fv] <- ""
}
```

### Initial Imputation Model Summary

**Total variables in dataset:** `r ncol(df_mice)`  
**Variables to be imputed:** `r sum(mice_dry$nmis > 0)`  
**Total missing values:** `r sum(mice_dry$nmis)`

**Missing data by variable:**
```{r missing-vars-table}
kable(miss_vars)
```

## Setup Predictor Matrix (strict temporal + fellow-eye logic)

```{r setup-predictors}
# Start with empty matrix, then opt in predictors explicitly
pred_matrix <- matrix(0, nrow = ncol(df_mice), ncol = ncol(df_mice))
colnames(pred_matrix) <- names(df_mice)
rownames(pred_matrix) <- names(df_mice)

# Never impute IDs (row = 0) and never use them as predictors (col = 0)
for (idv in c("row_id", "e_mrn_deidentified")) {
  if (idv %in% rownames(pred_matrix)) pred_matrix[idv, ] <- 0
  if (idv %in% colnames(pred_matrix)) pred_matrix[, idv] <- 0
}

# Never impute but can be used as predictors: ever_lapse_binary, cohort_id
if ("ever_lapse_binary" %in% rownames(pred_matrix)) pred_matrix["ever_lapse_binary", ] <- 0
if ("cohort_id" %in% rownames(pred_matrix)) pred_matrix["cohort_id", ] <- 0

# Fellow-eye and helper vars are predictors only (never imputed)
fellow_vars <- c("fellow_baseline_VA", "fellow_intermediate_VA", "fellow_outcome_VA", "fellow_dr", "n_eyes")
for (fv in fellow_vars) {
  if (fv %in% rownames(pred_matrix)) pred_matrix[fv, ] <- 0
}

# Variable groups
demographic_vars <- c("age", "gender_cat", "race_ethnic_cat", "insurance_cat")
global_comorbidity_vars <- c("CCI", "DCSI")
ocular_comorbidity_vars <- c("glaucoma_bef_hitplus_cat", "otherretina_bef_hitplus_cat", "catsurg_before_hitplus_cat")
treatment_vars <- c("anti_VEGF", "focal_laser_flag", "PRP_flag", "other_inject")

# baseline_VA_logMAR predictors (no future info)
if ("baseline_VA_logMAR" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 1
  for (v in global_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 1
  for (v in ocular_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", v] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "ever_lapse_binary"] <- 1
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["baseline_VA_logMAR", "fellow_baseline_VA"] <- 1
}

# intermediate_VA_logMAR predictors (no treatments, no future)
if ("intermediate_VA_logMAR" %in% rownames(pred_matrix)) {
  if ("baseline_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "baseline_VA_logMAR"] <- 1
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 1
  for (v in global_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 1
  for (v in ocular_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", v] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "ever_lapse_binary"] <- 1
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["intermediate_VA_logMAR", "fellow_baseline_VA"] <- 1
  # Explicitly exclude future/fellow-intermediate via omission (remain 0)
}

# outcome_VA_logMAR predictors (allow treatments and fellow baseline/outcome)
if ("outcome_VA_logMAR" %in% rownames(pred_matrix)) {
  if ("baseline_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "baseline_VA_logMAR"] <- 1
  if ("intermediate_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "intermediate_VA_logMAR"] <- 1
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  for (v in global_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  for (v in ocular_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "ever_lapse_binary"] <- 1
  for (v in treatment_vars) if (v %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", v] <- 1
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "fellow_baseline_VA"] <- 1
  if ("fellow_outcome_VA" %in% colnames(pred_matrix)) pred_matrix["outcome_VA_logMAR", "fellow_outcome_VA"] <- 1
}

# person_dr predictors (pre-outcome only, include fellow_baseline_VA and fellow_dr)
if ("person_dr" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  for (v in global_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  for (v in ocular_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  if ("baseline_VA_logMAR" %in% colnames(pred_matrix)) pred_matrix["person_dr", "baseline_VA_logMAR"] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["person_dr", "ever_lapse_binary"] <- 1
  for (v in treatment_vars) if (v %in% colnames(pred_matrix)) pred_matrix["person_dr", v] <- 1
  if ("fellow_baseline_VA" %in% colnames(pred_matrix)) pred_matrix["person_dr", "fellow_baseline_VA"] <- 1
  if ("fellow_dr" %in% colnames(pred_matrix)) pred_matrix["person_dr", "fellow_dr"] <- 1
}

# Demographics predict each other
for (v1 in demographic_vars) for (v2 in demographic_vars) {
  if (!identical(v1, v2) && v1 %in% rownames(pred_matrix) && v2 %in% colnames(pred_matrix)) pred_matrix[v1, v2] <- 1
}

# CCI and DCSI predict each other and are predicted by demographics and exposure
if ("CCI" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["CCI", v] <- 1
  if ("DCSI" %in% colnames(pred_matrix)) pred_matrix["CCI", "DCSI"] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["CCI", "ever_lapse_binary"] <- 1
}
if ("DCSI" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["DCSI", v] <- 1
  if ("CCI" %in% colnames(pred_matrix)) pred_matrix["DCSI", "CCI"] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["DCSI", "ever_lapse_binary"] <- 1
}

# Cataract surgery predicted by demographics/comorbidities/exposure
if ("catsurg_before_hitplus_cat" %in% rownames(pred_matrix)) {
  for (v in demographic_vars) if (v %in% colnames(pred_matrix)) pred_matrix["catsurg_before_hitplus_cat", v] <- 1
  for (v in global_comorbidity_vars) if (v %in% colnames(pred_matrix)) pred_matrix["catsurg_before_hitplus_cat", v] <- 1
  if ("ever_lapse_binary" %in% colnames(pred_matrix)) pred_matrix["catsurg_before_hitplus_cat", "ever_lapse_binary"] <- 1
}

# Treatment variables predict each other
for (v1 in treatment_vars) for (v2 in treatment_vars) {
  if (!identical(v1, v2) && v1 %in% rownames(pred_matrix) && v2 %in% colnames(pred_matrix)) pred_matrix[v1, v2] <- 1
}

log_message("Predictor matrix configured with strict temporal and fellow-eye constraints")

# Save the predictor matrix to CSV for inspection
pred_matrix_df <- as.data.frame(pred_matrix)
pred_matrix_df <- cbind(Variable = rownames(pred_matrix_df), pred_matrix_df)
write.csv(pred_matrix_df, file.path(reanalysis_data_dir, "predictor_matrix.csv"), row.names = FALSE)
log_message(paste0("Saved predictor matrix to: ", file.path(reanalysis_data_dir, "predictor_matrix.csv")))

# Simple verification prints
pred_counts <- data.frame(
  Variable = rownames(pred_matrix),
  N_Predictors = rowSums(pred_matrix > 0),
  N_Times_Used_As_Predictor = colSums(pred_matrix > 0)[rownames(pred_matrix)]
)
pred_counts <- pred_counts[order(-pred_counts$N_Predictors), ]
print(pred_counts)

# Detailed predictors for key variables
for (var in c("baseline_VA_logMAR", "intermediate_VA_logMAR", "outcome_VA_logMAR", "person_dr")) {
  if (var %in% rownames(pred_matrix)) {
    predictors <- colnames(pred_matrix)[pred_matrix[var, ] == 1]
    log_message(paste0("Predictors for ", var, " (", length(predictors), "): ", paste(predictors, collapse = ", ")))
  }
}
```

### Final Imputation Model Configuration

```{r final-model-config}
missing_info <- data.frame(
  Variable = names(mice_dry$nmis),
  N_Missing = mice_dry$nmis,
  stringsAsFactors = FALSE
)
